Supplementary Information: Vulnerability of tree growth

### 2025-03-21 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
MACROSCOPIC ORGANIZATION OF FRAMING:
- annotated paragraphs with comments summarizing content
- rearranged wording of title and main hypothesis  to have consistent ordering of drought characteristics, species effects, and microenvironment (applied that order)
- cleaned up a few old comments

**Title:** Tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments, and drought characteristics

@@

Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but it may be vulnerable to climate anomalies like drought [@tao_increasing_2022].  
- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- Climate anomalies like drought can lead to stand-level growth reductions, along with increased mortality [@bennett_sensitivity_2023; @bauman_tropical_2022].  
<!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
<!-- - Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017]. -->
- Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.  
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.

Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns.
- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).   
- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->

Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under hot/dry conditions.  
- evolutionary timescales - e.g. dry season deciduousness   
- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
- ecological timescales - e.g. distribution of trees in relation to wetness  
-  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.     
- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  
- Species with drought tolerant traits like deep roots, more negative turgor loss point and ... can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].  
- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years. 
- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.  
- Moreover, drought characteristics can positively or negatively interfere with these factors, leading to consistent or heterogenous sensitivities across drought events.

@@

ft <- flextable(sumtab)
ft <- set_header_labels(ft,
  na = ""
)
ft <- hline(ft, i = c(1, 4, 7), part = "body")
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
# ft <- height(ft, height = 0.3)
ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
ft <- width(ft, 1, width = 3.5)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
ft
```
<!---KAT framing---->

<!---High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
-Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity   -->

Here we examine individual tree growth during two known ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to disentangle the drivers of tree growth sensitivity to drought. 
We hypothesise that tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments and drought characteristics. 
<!-- We ask: i) How much do species vary in drought sensitivity of tree growth? What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought?  -->
Specifically, we expect that i) both species and individuals respond differently to different droughts. 
ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.

>>>>

**Title:** Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

@@

<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].  
- Climate anomalies like drought can lead to stand-level growth reductions, along with increased mortality [@bennett_sensitivity_2023; @bauman_tropical_2022].  
<!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
<!-- - Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017]. -->
- Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.  
- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns.
- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).   
- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->

<!--PARAGRAPHS ON HOW SPECIES TRAITS AND INDIVIDUAL CHARACTERISTICS SHAPE DROUGHT RESPONSES (SEPARATE SPECIES AND MICROENVIRONMENT)-->
Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under hot/dry conditions.  
- evolutionary timescales - e.g. dry season deciduousness   
- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
- ecological timescales - e.g. distribution of trees in relation to wetness  
-  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.     
- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  
- Species with drought tolerant traits like deep roots, more negative turgor loss point and ... can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].  
- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years. 
- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.  
- Moreover, drought characteristics can positively or negatively interfere with these factors, leading to consistent or heterogenous sensitivities across drought events.

@@

ft <- flextable(sumtab)
ft <- set_header_labels(ft,
  na = ""
)
ft <- hline(ft, i = c(1, 4, 7), part = "body")
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
# ft <- height(ft, height = 0.3)
ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
ft <- width(ft, 1, width = 3.5)
# ft <- width(ft, 2, width = 1.75)
# ft <- width(ft, 3, width = 3)
ft
```

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
Specifically, we test predictions that
i) both species and individuals respond differently to different droughts. 
ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
